---
title: "test"
author: "Dylan Tartarini, Mirco Balduini"
output:
  word_document: default
  html_notebook: default
---

>Introduction: 

This is a brief presentation regarding our study on electricity prices from the Italian market. Our goal is to compare approaches for modelling and forecasting complex, high-frequency data in a semi-automatic way. More details on why this issue is relevant and on the electricity prices univariate time series analysis can be found in this work's presentation freely available in our GitHub repos. 
We dispose of the hourly time series of the Unique National Price (PUN) starting from Jan, 1st, 2008 up to Dec, 31st, 2019. PUN is the reference price detected on Italian Power Exchange (IPEX): data are free to use and can be found on the web-page of the Italian Regulator for the Energy Markets (GME) following this link: https://www.mercatoelettrico.org/it/.  
Let's dive in!

>Set-Up  

Loading needed packages..
```{r}
library(urca)
library(rmarkdown)
library(stats)
library(tseries)
library(tidyr)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(lubridate)
library(datetime)
library(colortools)
library(zoo)
library(gridExtra)
library(astsa)
library(describedata)
library(FactoMineR)
library(factoextra)
library(data.table)
library(ARDL)
library(dLagM)
library(prophet)
library(forecast)
```

Setting directory and loading data..

```{r}
PUN_db <- read_excel("C:/Users/Utente/Google Drive/PUN_TS analysis/PUN_TS analysis/Analisi ts PUN/PUN_2008_2019.xlsx")

PUN_db <- read_excel("PUN_2008_2019.xlsx", sheet = "2008_2019")

data.frame(PUN_db)
```

>Data Manipulation

Manipulation of the full hourly dataset..
```{r}
PUN_db <- PUN_db %>% 
  separate(Hour, into = c("Date", "Time"), sep = " ", remove = FALSE) %>%
  mutate(Date = lubridate::as_date(Date, format = "%Y/%m/%d"),
         Time = hms::as_hms(str_c(Time, ":00")))

PUN_db$Date <- PUN_db$Hour <- NULL

PUN_db$DateReal <- as.Date(PUN_db$DateReal, "%Y/%m/%d")

PUN_db$DateTime <- as.POSIXct(paste(PUN_db$DateReal, PUN_db$Time), format = "%Y-%m-%d %H:%M:%S")

PUN_db <- unique(setDT(PUN_db)[order(DateTime, PUN)], by = "DateTime")

```

Widening data with the purpose of summarizing with daily averages..

```{r}
PUN_db_wide <- pivot_wider(data = PUN_db, id_cols = DateReal, names_from = Time, values_from = PUN)

PUN_db_wide <- data.frame(PUN_db_wide)

colnames(PUN_db_wide)[2:25] <- c("h24","h01","h02", "h03", "h04", "h05", "h06", "h07",
                                   "h08", "h09", "h10", "h11", "h12", "h13", "h14",
                                   "h15", "h16", "h17", "h18","h19","h20", "h21",
                                   "h22","h23")
PUN_db_wide <- PUN_db_wide[,c("DateReal", "h01","h02", "h03", "h04", "h05", "h06", "h07",
                                  "h08", "h09", "h10", "h11", "h12", "h13", "h14",
                                  "h15", "h16", "h17", "h18","h19","h20", "h21",
                                  "h22","h23", "h24")]

PUN_db_wide$DailyPun <- rowMeans(PUN_db_wide[2:25])

PUN_db_wide$DailyPun[is.na(PUN_db_wide$DailyPun)] <- mean(PUN_db_wide$DailyPun, na.rm = TRUE)   

for(i in 1:ncol(PUN_db_wide[2:26])){
  PUN_db_wide[is.na(PUN_db_wide[,i]), i] <- mean(PUN_db_wide[,i], na.rm = TRUE)               
}
```

Creating other dates variables..

```{r}
  
PUN_db$Week <- as.Date(cut(PUN_db$DateReal, breaks = "week"))  #start.on.monday = FALSE)) # changes weekly                                                                break point to Sunday
PUN_db$WeekNumber <- isoweek(ymd(PUN_db$DateReal))

PUN_db$Month<- month(PUN_db$DateReal)

PUN_db$MonthName <- month.abb[PUN_db$Month] 
                              
PUN_db$Year <- year(PUN_db$DateReal)

PUN_db$DayCount <- day(PUN_db$DateReal)

PUN_db$YearMonth <- as.yearmon(PUN_db$DateReal)

PUN_db$DayName <- strftime(PUN_db$DateReal, "%A")

PUN_db_wide$Year <- year(PUN_db_wide$DateReal)
```

Looking our 2 dataframes..

```{r}
head(PUN_db)
```

```{r}
head(PUN_db_wide)
```

Computing averages and standard deviations 

```{r}
mean <- mean(PUN_db$PUN)

hourly_average <- PUN_db %>%
  group_by(Time) %>%
  summarize(mean_PUN = mean(PUN))


daily_average <- PUN_db %>%
  group_by(DateReal) %>%
  summarize(mean_PUN = mean(PUN))

daily_average$DayOfYear <- strftime(daily_average$DateReal, "%j")
daily_average$Year <- year(daily_average$DateReal)
daily_average$Year <- as.factor(daily_average$Year)

weekly_average <- PUN_db %>%
  group_by(Week) %>%
  summarize(mean_PUN = mean(PUN))

weekly_average$WeekNumber <- isoweek(ymd(weekly_average$Week))
weekly_average$Year <- year(weekly_average$Week)
weekly_average$Year <- as.factor(weekly_average$Year)
weekly_average$WeekNumber <- as.factor(weekly_average$WeekNumber)

mon_avg_perYear <- aggregate(PUN_db$PUN, list(Date=format(PUN_db$YearMonth, "%Y-%m")),mean)
colnames(mon_avg_perYear) <- c("Date", "avg_PUN")
mon_avg_perYear <- mon_avg_perYear %>% 
  separate(Date, into = c("Year", "Month"), sep = "-", remove = FALSE)
mon_avg_perYear$Year <- as.factor(mon_avg_perYear$Year)
mon_avg_perYear$Month <- as.factor(mon_avg_perYear$Month)


yearly_average <- PUN_db %>%
  group_by(Year) %>%
  summarize(mean_PUN = mean(PUN))

sd <- sd(PUN_db$PUN)

sd_hourly <- PUN_db %>%
  group_by(Time) %>%
  summarize(sd_PUN = sd(PUN))

sd_yearly <- PUN_db %>%
  group_by(Year) %>%
  summarize(sd_PUN = sd(PUN))

sd_montly_perYear <- PUN_db %>%
  group_by(YearMonth) %>%
  summarize(sd_PUN = sd(PUN))

```

From Data Frames to time series..

```{r}

dec19 <- subset(PUN_db, (Year == 2019 & MonthName == "Dec"))

start_date <- c(2008,1,1)
PUN_ts <- ts(PUN_db$PUN, start = start_date, frequency = 24*365.25)

dailyPUN_ts <- ts(PUN_db_wide$DailyPun, start = start_date, frequency = 365.25)

PUN_2018 <- subset(PUN_db, Year == 2018)

hourlyPun_2018 <- ts(PUN_2018$PUN, start = c(2018, 1, 1) ,frequency = 24*365)

PUN_2018_19 <- subset(PUN_db, Year >= 2018)

hourlyPUN <- ts(PUN_2018_19$PUN, start = c(2018, 1, 1) ,frequency = 24*365)

dec19_ts <- ts(dec19$PUN, start = c(2019,12,01), freq = 24*31)
```

> Exploratory Data Analysis 

The full hourly dataset 

```{r}

ggplot(PUN_db, aes(x=DateTime, y=PUN))+
  geom_line(color="mediumblue")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Hourly Price of Electricity")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))
```
Same dataset, but for daily average observations
```{r}
ggplot(daily_average, aes(x=DateReal, y=mean_PUN))+
  geom_line(color = "mediumblue")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Average Price of Electricity per day")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```
Distribution of PUN across hours of the day

```{r}
ggplot(PUN_db, aes(x=Time, y=PUN))+
  geom_boxplot(aes(group=Time), color = "blue", fill = "blue", alpha = 0.2, 
               outlier.color = "red", outlier.fill = "red", outlier.size = 3)+
  xlab("Hours")+
  ylab("Price (EUR/MWh)")+
  ggtitle("Distribution of Hourly Prices")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))
```

Distribution of PUN across months of the year

```{r}
as.factor(PUN_db$MonthName)

ggplot(PUN_db, aes(x=MonthName, y=PUN))+
  geom_boxplot(aes(group=Month), color = "blue", fill = "blue", alpha = 0.2, 
               outlier.color = "red", outlier.fill = "red", outlier.size = 3)+
  scale_x_discrete(limits = month.abb)+
  xlab("Months")+
  ylab("Price (EUR/MWh)")+
  ggtitle("Distribution of Monthly Average Prices")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))
```
Distribution of PUN across weeks of the year

```{r}
ggplot(PUN_db, aes(x=WeekNumber, y=PUN))+
  geom_boxplot(aes(group=WeekNumber), color = "blue", fill = "blue", alpha = 0.2, 
               outlier.color = "red", outlier.fill = "red", outlier.size = 3)+
  xlab("Week Number")+
  ylab("Price (EUR/MWh)")+
  ggtitle("Distribution of Weekly prices")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))
```

Distribution of PUN across days of the week

```{r}
PUN_db$DayName <- factor(PUN_db$DayName, c("domenica", "lunedì", "martedì", "mercoledì",
                                           "giovedì", "venerdì", "sabato"))

ggplot(PUN_db, aes(x=DayName, y=PUN))+
  geom_boxplot(aes(group=DayName), color = "blue", fill = "blue", alpha = 0.2, 
               outlier.color = "red", outlier.fill = "red", outlier.size = 3)+
  xlab("Day of Week")+
  ylab("Price (EUR/MWh)")+
  ggtitle("Distribution of prices during days of week")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))
```

>EDA: Visualizing Seasonality

```{r}
year_pal1 <- sequential(color = "red", percentage = 5, what = "value")

year_pal2 <- sequential(color = "green", percentage = 5, what = "value")

year_pal3 <- sequential(color = "cyan3", percentage = 5, what = "value")
```
The next three plots show that electricity prices follow the same pattern during different years

```{r}
ggplot(mon_avg_perYear, aes(x = Month, y = avg_PUN, group = Year))+
  geom_line(aes(colour = Year))+
  ggtitle("avg Monthly Prices Comparison Between Years")+
  ylab("avg monthly PUN (EUR/MWh)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_color_manual(values = year_pal1)
```

```{r}
ggplot(weekly_average, aes(x = WeekNumber, y = mean_PUN, group = Year))+
  geom_line(aes(colour = Year))+
  ggtitle("avg Weekly Prices Comparison Between Years")+
  ylab("avg weekly PUN (EUR/MWh)")+
  scale_x_discrete(limits = c(1:52), breaks =c(1,6,12,17,22,27,32,37,42,47,52))+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_color_manual(values = year_pal2)
```

```{r}
ggplot(daily_average, aes(x = DayOfYear, y = mean_PUN, group = Year))+   
  geom_line(aes(colour = Year))+
  ggtitle("avg Daily Prices Comparison Between Years")+
  ylab("avg daily PUN (EUR/MWh)")+
  scale_x_discrete(limits = c(1:365), breaks =c(1,30,90, 150, 210,270, 330, 365))+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_color_manual(values = year_pal3)
```

Electricity Prices show a **common structure** both in the conditional mean..
```{r}
plt_h2 <- ggplot(PUN_db_wide , aes(x=DateReal, y=h02))+                                                   
  geom_line(color="blue")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Price of Electricity (2 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_h9 <- ggplot(PUN_db_wide , aes(x=DateReal, y=h09))+
  geom_line(color="green")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Price of Electricity (9 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_h11 <- ggplot(PUN_db_wide , aes(x=DateReal, y=h11))+
  geom_line(color="red")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Price of Electricity (11 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_h17 <- ggplot(PUN_db_wide , aes(x=DateReal, y=h17))+
  geom_line(color="orange")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Price of Electricity (5 PM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_h19 <- ggplot(PUN_db_wide , aes(x=DateReal, y=h19))+
  geom_line(color="cyan3")+
  xlab("Years")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Price of Electricity (7 PM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plt_h2, plt_h9, plt_h11, plt_h17, plt_h19, nrow = 5)
```
..and in the conditional conditional variance

```{r}
PUN_db_wide_h <- data.frame(PUN_db_wide$DateReal, PUN_db_wide$h02, PUN_db_wide$h09, PUN_db_wide$h11, 
                             PUN_db_wide$h17, PUN_db_wide$h19 )

colnames(PUN_db_wide_h)[1:6] <- c("DateReal","h02", "h09", "h11", "h17","h19")

mean_h2 <- mean(PUN_db_wide_h$h02)

mean_h9 <- mean(PUN_db_wide_h$h09)

mean_h11 <- mean(PUN_db_wide_h$h11)

mean_h17 <- mean(PUN_db_wide_h$h17)

mean_h19 <- mean(PUN_db_wide_h$h19)

PUN_db_wide_h$Dev2 <- abs(PUN_db_wide_h$h02 - mean_h2)

PUN_db_wide_h$Dev9 <- abs(PUN_db_wide_h$h09 - mean_h9)

PUN_db_wide_h$Dev11 <- abs(PUN_db_wide_h$h11 - mean_h11)

PUN_db_wide_h$Dev17 <- abs(PUN_db_wide_h$h17 - mean_h17)

PUN_db_wide_h$Dev19 <- abs(PUN_db_wide_h$h19 - mean_h19)

plt_Dev2 <- ggplot(PUN_db_wide_h , aes(x = DateReal, y = Dev2))+      
  geom_line(color="blue")+
  xlab("Years")+
  ylab("Dev")+
  ggtitle("Deviation from the Mean Price (2 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "top")

plt_Dev9 <- ggplot(PUN_db_wide_h , aes(x=DateReal, y=Dev9))+
  geom_line(color="green")+
  xlab("Years")+
  ylab("Dev")+
  ggtitle("Deviation from the Mean Price (9 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_Dev11 <- ggplot(PUN_db_wide_h , aes(x=DateReal, y=Dev11))+
  geom_line(color="red")+
  xlab("Years")+
  ylab("Dev")+
  ggtitle("Deviation from the Mean Price (11 AM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_Dev17 <- ggplot(PUN_db_wide_h , aes(x=DateReal, y=Dev17))+
  geom_line(color="orange")+
  xlab("Years")+
  ylab("Dev")+
  ggtitle("Deviation from the Mean Price (5 PM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

plt_Dev19 <- ggplot(PUN_db_wide_h , aes(x=DateReal, y=Dev19))+
  geom_line(color="cyan3")+
  xlab("Years")+
  ylab("Dev")+
  ggtitle("Deviation from the Mean Price (7 PM)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plt_Dev2, plt_Dev9, plt_Dev11, plt_Dev17, plt_Dev19, nrow = 5)

```
> EDA: Stationarity Check of the Hourly Series

```{r}
pp.test(PUN_ts, alternative = c("s"),
        type = c("Z(alpha)", "Z(t_alpha)"), lshort = TRUE)  #p-value is 0.01, so there is no unit root                                                                                    -> PUN is stationary
```
```{r}
summary(ur.kpss(PUN_ts)) #pun is not stationary according to KPSS test 1

summary(ur.kpss(PUN_ts, lags = "short"))   #pun is not stationary according to KPSS test 2

kpss.test(PUN_ts, null = "Level", lshort = TRUE) #pun is not stationary according to KPSS test 3

```
```{r}                 
                 #this is the standard ADF test from the tseries() package     
adf.test(PUN_ts) #The p-value is smaller than 0.05 --> 
                 #We can reject the null hypothesis. The null hypothesis is that the data are                               non-stationary.

                                         #ADF test from the urca() package for having control on the lags and type of ts
ur.df(PUN_ts, type = "none", lags = 24)  #value of t-stat is: -4.1668 --> p-value is less that 0.01 -->
                                         #we can reject the null hypothesis
```

```{r}
acf(PUN_ts, lag.max=24*2)      #with seasonal patterns (which electricity does have),
pacf(PUN_ts, lag.max = 24*2)   #order of lag is 2*m, where m is the periodicity  
```
```{r}
dec_pun <- decompose(PUN_ts, type="mult")  #decomposition of elements of the ts of electricity prices..
plot(dec_pun)

```
>EDA: Stationarity Check on Daily AVG TS

```{r}
pp.test(dailyPUN_ts, alternative = c("s"),
        type = c("Z(alpha)", "Z(t_alpha)"), lshort = TRUE)  #p-value is 0.01, so there is no unit root -> 
                                                            #dailyPUN_ts is stationary
```

```{r}
summary(ur.kpss(dailyPUN_ts)) #dailyPUN_ts is not stationary according to KPSS test 
```
```{r}
                      #this is the standard ADF test from the tseries() package     
adf.test(dailyPUN_ts) #The p-value is smaller than 0.05 --> 
                      #We can reject the null hypothesis. The null hypothesis is that the data are                               non-stationary.

                                              #ADF test from the urca() package for having control on the                                               lags and type of ts
ur.df(dailyPUN_ts, type = "none", lags = 12)  #value of t-stat is: -1.4064  --> p-value is less that 0.01                                               --> we can reject the null hypothesis
```
```{r}
acf(dailyPUN_ts, lag.max=30)      #with seasonal patterns (which electricity does have),
pacf(dailyPUN_ts, lag.max = 30)   #order of lag is 2*m, where m is the periodicity  
```
```{r}
dec_pun_daily <- decompose(dailyPUN_ts, type="mult")  #decomposition of elements of the ts of electricity                                                        prices..
plot(dec_pun_daily)

```

> NOWCASTING: sARIMA on the HOURLY SUBSET 

We first run *auto.arima()* and then adapt a seasonal pattern employing a minimum of domain knoledge.  
Looking at the hourly subset employed for modelling:
```{r}
ts.plot(dec19_ts)
```

Splitting the series in training and test set

```{r}
dec19_train <- head(dec19_ts, round(length(dec19_ts) * 0.9))

h <- length(dec19_ts) - length(dec19_train)          

dec19_test <- tail(dec19_ts, h)

autoplot(dec19_train) + autolayer(dec19_test)
```

Running *auto.arima()* suggest to use **ARIMA(5,1,2)** as a starting point..

```{r}
auto.arima(dec19_train)  #suggested: ARIMA(5,1,2)
```

Following auto.arima's suggestion and adding the concept of seasonality: looking at the model..
```{r}
sARIMA_dec19.train <- sarima(dec19_train, p = 5, d = 1, q = 2, P = 1, D = 0, Q = 1, S = 24, details = TRUE, xreg = NULL, 
                              Model = TRUE, fixed = NULL, no.constant = TRUE)   #look at residuals 
```

..and forecasting

```{r}
fcst_dec19.sar <- sarima.for(dec19_train, n.ahead = 74 , p = 5, d = 1, q = 2, P = 1, D = 0, Q = 1, S = 24,
                              xreg = NULL, plot = TRUE, plot.all = FALSE, fixed = NULL, no.constant = TRUE)  
```
Plotting sARIMA's forecast against actual values: the result seem satisfying.

```{r}
dec19_df_conf <- slice(dec19, 671:744)

dec19_df_conf$Pred <- cbind(fcst_dec19.sar$pred)

ggplot() + 
  geom_line(data = dec19, aes(x = DateTime, y = PUN)) + 
  geom_line(data = dec19_df_conf, aes(x = DateTime , y = Pred), color='red')+
  theme_light()+
  xlab("Time")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Hourly Price of Electricity (sARIMA)")+
  theme(plot.title = element_text(hjust = 0.5))   
```
Computing **Performance Metrics** for sARIMA model..

```{r}
sARIMA_dec19 <- sarima(dec19_ts, p = 5, d = 1, q = 2, P = 1, D = 0, Q = 1, S = 24, details = FALSE, xreg = NULL, 
                       Model = FALSE, fixed = NULL, no.constant = TRUE) #model on the whole december series

MAE_sARIMA_hourly <- mean(abs(residuals(sARIMA_dec19$fit)))
MAPE_sARIMA_hourly <- mean(100*abs(residuals(sARIMA_dec19$fit))/dec19_ts)
RMSE_sARIMA_hourly <- sqrt(mean(residuals(sARIMA_dec19$fit)^2))

```

> NOWCASTING: Prophet on the HOURLY SUBSET

We can now run the *prophet* package on the same series and verify which method performs better. 

```{r}
dec_19_df <- dec19[1:672,]

HourlyPUN_df <- data.frame(dec_19_df$DateTime, dec_19_df$PUN)

colnames(HourlyPUN_df) <- c("ds", "y")   #Prophet need exactly those two names for variables

mod_prophet_hourly <- prophet(df = HourlyPUN_df, growth = "linear", yearly.seasonality = FALSE, weekly.seasonality = TRUE, 
                              daily.seasonality = TRUE, seasonality.mode = "additive", fit = TRUE ) #Prophet model
```

Prophet method needs a dataframe with future dates for storing Prophet's predictions. After creating it, we are able to obtain predictions from the model

```{r}
future_dec <- make_future_dataframe(mod_prophet_hourly, periods = 24*3, freq = 3600)  

fcst_hourly_prophet <- predict(mod_prophet_hourly, future_dec)

plot(mod_prophet_hourly, fcst_hourly_prophet)
```

We can also plot the components of the model 

```{r}
prophet_plot_components(mod_prophet_hourly, fcst_hourly_prophet) 
```
Prophet method provide a useful function for **Cross validation**:

```{r}
prophet_cv_hourly <- cross_validation(mod_prophet_hourly, initial = length(HourlyPUN_df)*0.8, horizon = 24*3, units = 'hours')

head(prophet_cv_hourly)
```
```{r}
plot_cross_validation_metric(prophet_cv_hourly, "mae")
```
```{r}
plot_cross_validation_metric(prophet_cv_hourly, "rmse")
```
```{r}
plot_cross_validation_metric(prophet_cv_hourly, "mape")
```
Prophet also permits to output the performance metrics from its cross validation method: 

```{r}
prophet_metrics_h <- performance_metrics(prophet_cv_hourly)

head(prophet_metrics_h)

MAPE_prophet_h <- mean(prophet_metrics_h$mape)
MAE_prophet_h <- mean(prophet_metrics_h$mae)            
RMSE_prophet_h <-  mean(prophet_metrics_h$rmse)
```

Actual Values and Fitted values, with confidence intervals. 

```{r}
ggplot()+ 
  geom_line(data = dec19, aes(x = DateTime, y = PUN))+ 
  geom_line(data = fcst_hourly_prophet, aes(x = ds, y = yhat), color='cyan3')+
  geom_line(data = fcst_hourly_prophet, aes(x = ds, y = yhat_lower), color='mediumblue', alpha=0.5)+
  geom_line(data = fcst_hourly_prophet, aes(x = ds, y = yhat_upper), color='mediumblue', alpha=0.5)+
  theme_light()+
  xlab("Time")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Hourly Price of Electricity (Prophet)")+
  theme(plot.title = element_text(hjust = 0.5))
```

> NOWCASTING: PROPHET vs sARIMA

```{r}
fcst_hourly_prophet_sub <- slice(fcst_hourly_prophet, 671:744)

dec19_df_conf$Pred_Prophet <- fcst_hourly_prophet_sub$yhat

ggplot(data = dec19_df_conf, aes(x = DateTime)) +
  geom_line(aes(y = PUN, colour = "Actual")) +
  geom_line(aes(y = Pred_Prophet, colour = "Prophet")) +
  geom_line(aes(y = Pred, colour = "sARIMA")) +
  scale_colour_manual("", 
                      breaks = c("Actual", "Prophet", "sARIMA"),
                      values = c("black", "mediumblue", "red")) +
  xlab("Time") +
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Hourly Price of Electricity (Prophet VS sARIMA)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```
As can be seen by the figure, we can obtain similar results between Prophet and sARIMA models; what gives the edge to Prophet however, is the extreme user-friendliesness and time saving capabilities: sARIMA has to be tuned manually and necessitates of way more domain knowledge than Prophet does.

> LONG RUN FORECASTING: sARIMA on daily data

Since sARIMA and Prophet both perform well in the task of forecasting day-ahead hourly prices, we try to test them with the task of forecasting daily prices one year ahead. We start with sARIMA.  

Splitting in training and test set..

```{r}
dailyPUN_train <- subset(PUN_db_wide, Year <= 2018)

dailyPUN_test <- subset(PUN_db_wide, Year > 2018)

dailyPUN_train_ts <- ts(dailyPUN_train$DailyPun, start = c(2008,1,1), freq = 365.25)

dailyPUN_test_ts <- ts(dailyPUN_test$DailyPun,  start = c(2019,1,1), freq = 365)

autoplot(dailyPUN_train_ts) + autolayer(dailyPUN_test_ts)
```
..and using *auto.arima()* on the training set.

```{r}
auto.arima(dailyPUN_train_ts, seasonal = TRUE, stepwise = TRUE)
```

We run the suggested model on the set, adding some seasonality to it:

```{r}
sARIMA_daily.train <- sarima(dailyPUN_train_ts, p = 5, d = 1, q = 1, P = 1, D = 0, Q = 1, S = 7, details = TRUE, xreg = NULL, 
                             Model = TRUE, fixed = NULL, no.constant = TRUE)
```
Long run forecasting with daily data. We already got the impression of something being off about long-run daily predictions of PUN.. 

```{r}
fcst_daily.test <- sarima.for(dailyPUN_train_ts, n.ahead = 365 , p = 5, d = 1, q = 1, P = 1, D = 0, Q = 1, S = 7,
                         xreg = NULL, plot = TRUE, plot.all = FALSE, fixed = NULL, no.constant = TRUE) 
```

```{r}
fcst_daily.test_df <- data.frame(fcst_daily.test$pred)

fcst_daily.test_df$Date <- seq(ymd("2019-1-1"), ymd("2019-12-31"), by = "days")

colnames(fcst_daily.test_df) <- c("Pred_PUN", "Date")

ggplot() + 
  geom_line(data = subset(PUN_db_wide, Year >= 2015), aes(x = DateReal, y = DailyPun)) + 
  geom_line(data = fcst_daily.test_df, aes(x = Date , y = Pred_PUN), color='red')+
  theme_light()+
  xlab("Time")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (sARIMA)")+
  theme(plot.title = element_text(hjust = 0.5))   
```
sARIMA modelling applied at the whole daily series as it is does not provide us with acceptable results. We store the Performance Metrics and then go on testing Prophet against the same series.

```{r}
sARIMA_daily <- sarima(dailyPUN_ts, p = 5, d = 1, q = 1, P = 1, D = 0, Q = 1, S = 7, details = FALSE, xreg = NULL, 
                             Model = FALSE, fixed = NULL, no.constant = TRUE)

MAE_sARIMA_daily <- mean(abs(residuals(sARIMA_daily$fit)))                 
MAPE_sARIMA_daily <- mean(100*abs(residuals(sARIMA_daily$fit))/dailyPUN_ts)
RMSE_sARIMA_daily <- sqrt(mean(residuals(sARIMA_daily$fit)^2))
```

From previous figure it should be obvious that sARIMA modelling is not enough for forecasting 365 days ahead, because of the non-linearities the ARIMA model class cannot identify. Will Prophet perform any better?

> LONG RUN FORECASTING: PROPHET on daily data

```{r}
DailyPUN_train_df <- data.frame(dailyPUN_train$DateReal, dailyPUN_train$DailyPun)

colnames(DailyPUN_train_df) <- c("ds", "y")

mod_prophet_daily <- prophet(df = DailyPUN_train_df, growth = "linear", yearly.seasonality = TRUE, weekly.seasonality = TRUE, 
                       daily.seasonality = "auto", holidays = NULL, seasonality.mode = "additive", fit = TRUE )
```

```{r}
future_d <- make_future_dataframe(mod_prophet_daily, periods = 365, freq = "day")

fcst_daily_prophet <- predict(mod_prophet_daily, future_d)

plot(mod_prophet_daily, fcst_daily_prophet)
```

```{r}
prophet_plot_components(mod_prophet_daily, fcst_daily_prophet) 
```

```{r}
prophet_cv_daily <- cross_validation(mod_prophet_daily, initial = length(DailyPUN_train_df)*0.8, horizon = 365.25, units = 'days')

head(prophet_cv_daily)
```

```{r}
plot_cross_validation_metric(prophet_cv_daily, "mae")
```

```{r}
plot_cross_validation_metric(prophet_cv_daily, "rmse")
```

```{r}
plot_cross_validation_metric(prophet_cv_daily, "mape")
```
```{r}
prophet_metrics_d <- performance_metrics(prophet_cv_daily)

head(prophet_metrics_d)

MAE_prophet_d <- mean(prophet_metrics_d$mae)
MAPE_prophet_d <- mean(prophet_metrics_d$mape)  
RMSE_prophet_d <- mean(prophet_metrics_d$rmse)
```

```{r}
fcst_daily_prophet$Year <- year(fcst_daily_prophet$ds) 

fcst_daily_prophet$ds <- as.Date(fcst_daily_prophet$ds)

ggplot() + 
  geom_line(data = subset(PUN_db_wide, Year >= 2015), aes(x = DateReal, y = DailyPun)) + 
  geom_line(data = subset(fcst_daily_prophet, Year >= 2015), aes(x = ds, y = yhat), color='cyan3')+
  geom_line(data = subset(fcst_daily_prophet, Year >= 2015), aes(x = ds, y = yhat_lower), color='blue', alpha=0.3)+
  geom_line(data = subset(fcst_daily_prophet, Year >= 2015), aes(x = ds, y = yhat_upper), color='blue', alpha=0.3)+
  theme_light()+
  xlab("Time")+
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (Prophet)")+
  theme(plot.title = element_text(hjust = 0.5))
```
Prophet performs better than sARIMA on long-run forecasting in terms of evaluation metrics but, both in additive and multiplicative seasonality mode, fails to forecast electricity prices as it should.

```{r}
dailyPUN_test <- data.frame(dailyPUN_test$DateReal, dailyPUN_test$DailyPun)

colnames(dailyPUN_test) <- c("DateReal", "DailyPun")

dailyPUN_test$Pred_sar <- fcst_daily.test_df$Pred_PUN

prophet_2019_daily <- subset(fcst_daily_prophet, Year >= 2019)

dailyPUN_test$Pred_Prophet <- prophet_2019_daily$yhat

ggplot(data = dailyPUN_test, aes(x = DateReal)) +
  geom_line(aes(y = DailyPun, colour = "Actual")) +
  geom_line(aes(y = Pred_Prophet, colour = "Prophet")) +
  geom_line(aes(y = Pred_sar, colour = "sARIMA")) +
  scale_colour_manual("", 
                      breaks = c("Actual", "Prophet", "sARIMA"),
                      values = c("black", "mediumblue", "red")) +
  xlab("Time") +
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (Prophet VS sARIMA)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```
>Working around the Long-Casting task: for loops and sliding windows 

One way we figured out to address the long-casting task and obtain believable predictions is to use a sliding window of one month of daily observations and running our models through it, forecasting three days ahead, dropping the first three observations and repeating. This comes at cost of losing a degree of automation and increasing complexity a little but, as the reader will notice, our strategy is worth exploring.

```{r}
db_2018 <- subset(PUN_db_wide, Year == 2018)

db_2018 <- data.frame(db_2018$DateReal, db_2018$DailyPun)

colnames(db_2018) <- c("DateReal", "DailyPun")

db_2018$month <- month(db_2018$DateReal)

dec18 <- subset(db_2018, month == 12)

row.names(dec18) <- NULL

db_2019 <- subset(PUN_db_wide, Year == 2019)

db_2019 <- data.frame(db_2019$DateReal, db_2019$DailyPun)

colnames(db_2019) <- c("DateReal", "DailyPun")

db_2019$month <- month(db_2019$DateReal) 

row.names(db_2019) <- NULL

daily_set <- rbind(dec18, db_2019)

daily_set <- daily_set[1:2]
```

**Prophet on a sliding window **  
We first initialize a counter and the needed data frames in which we'll store the new predictions. Then we employ a while loop to recursively run prophet. The results can be observed in the graph below.

```{r}
c <- 0
fcst_test_prophet <- data.frame()
running <- data.frame()
pred_running <- data.frame()


while(c < nrow(daily_set) - 33){    
  
  sample_p <- daily_set[(1+c):(30+c), ]
  
  colnames(sample_p) <- c("ds", "y")
  
  mod_prophet_test <- prophet(df = sample_p, growth = "linear", yearly.seasonality = "auto", weekly.seasonality = TRUE, 
                               daily.seasonality = "auto", holidays = NULL, seasonality.mode = "additive", fit = TRUE )
  
  future_test <- make_future_dataframe(mod_prophet_test, periods = 3, freq = "day")
  
  fcst_test_prophet <- predict(mod_prophet_test, future_test) 
  
  running <- fcst_test_prophet[31:33, c("ds", "yhat")] #these are the three rolling predictions..
  
  pred_running <- rbind(pred_running, running)  
  
  c <- c+3
}

daily_set <- daily_set[31:nrow(daily_set), ]

row.names(daily_set) <- NULL

daily_set <- daily_set[1:363, ]

pred_running <- cbind(pred_running, daily_set$DailyPun)

colnames(pred_running) <- c("date", "prophet_pred", "DailyPUN")

ggplot(data = pred_running, aes(x = date)) +
  geom_line(aes(y = DailyPUN, colour = "Actual")) +
  geom_line(aes(y = prophet_pred, colour = "PROPHET")) +
  scale_colour_manual("", 
                      breaks = c("Actual", "PROPHET"),
                      values = c("black", "mediumblue")) +
  xlab("Time") +
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (Prophet on sliding window)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```

As can be seen, our method works fairly well for prophet. We store the Performance Metrics from the predictions and then procede with a similar method for sARIMA modelling. 

```{r}
PROPHET_roll_acc <- accuracy(object = pred_running$prophet_pred, x = pred_running$DailyPUN, test = NULL)
```

**sARIMA on a sliding window of observations**  
Preparing the subset to run through with sARIMA: here, we don't have the chance to use *auto.arima()* and then decide what to do. Our approach is then to utilize a simple ARIMA(1,1,1)x(1,0,1)[7] since we are running through 31 observations at a time.  
Once again, we initialize the ausiliary data frames and then run the for loop, obtaining satisfactory results for our Long-casting purposes.

```{r}
daily_set_sar <- rbind(dec18, db_2019)

daily_set_sar <- daily_set_sar[1:2]

sample_sar <- data.frame()
predictions_sar <- data.frame()

for (i in seq(1, nrow(daily_set_sar)-30, 3)){    #1/3 of actual values
  
  sample_sar <- daily_set_sar$DailyPun[i:c(29+i)]
  
  dates <- daily_set_sar$DateReal[i:c(29+i)]
  
  sample_ts <-as.ts(sample_sar, start_date = dates[1], freq = 365)
  
  sARIMA <- arima(sample_ts, order=c(1,1,1), seasonal=list(order=c(1,0,1), period=7), include.mean = FALSE, method="ML")
  
  sARIMA_fcst <- forecast(sARIMA, h =3)
  
  predictions_sar <- rbind(predictions_sar, as.data.frame(as.numeric(sARIMA_fcst$mean)))
}

daily_set_sar <- daily_set_sar[31:nrow(daily_set_sar), ]

row.names(daily_set_sar) <- NULL

daily_set_sar <- cbind(daily_set_sar, predictions_sar)

colnames(daily_set_sar) <- c("date", "daily_PUN", "predictions_sar")

ggplot(data = daily_set_sar, aes(x = date)) +
  geom_line(aes(y = daily_PUN, colour = "Actual")) +
  geom_line(aes(y = predictions_sar, colour = "sARIMA")) +
  scale_colour_manual("", 
                      breaks = c("Actual", "sARIMA"),
                      values = c("black", "red")) +
  xlab("Time") +
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (sARIMA on sliding window)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```

As with Prophet, we can store the resulting Performance Metrics..

```{r}
sARIMA_roll_acc <- accuracy(object = daily_set_sar$predictions_sar, x = daily_set_sar$daily_PUN, test = NULL)
```

.. and then plot the predictions obtained from the two models.

```{r}

daily_set_sar <- daily_set_sar[1:(nrow(daily_set_sar)-3), ]

daily_df_fin <- cbind(daily_set_sar, pred_running$prophet_pred)

colnames(daily_df_fin) <- c("date", "DailyPUN", "sARIMA_pred", "Prophet_pred")

ggplot(data = daily_df_fin, aes(x = date)) +
  geom_line(aes(y = DailyPUN, colour = "Actual")) +
  geom_line(aes(y = sARIMA_pred, colour = "sARIMA")) +
  geom_line(aes(y = Prophet_pred, colour = "PROPHET")) +
  scale_colour_manual("", 
                      breaks = c("Actual", "sARIMA", "PROPHET"),
                      values = c("black", "red", "mediumblue")) +
  xlab("Time") +
  ylab("PUN (EUR/MWh)")+
  ggtitle("Predicted Daily Price of Electricity (sARIMA vs PROPHET on sliding window)")+
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```
Our job here is done, as both sARIMA and Prophet are now able to forecast with a satisfactory degree of accuracy. sARIMA is slightly better according to MAPE, but Prophet as probably the edge here, because of the lesser effort it requires, in terms both of learning curve and looping automatization. 

Thanks for reading us! 





